How and Why to run a Hack Week

### 2017-09-30 sections/design.tex

A hackweek requires a flexible workspace environment with ample opportunity for re-configuration. 
Participants must have access to rooms that can accommodate lectures combined with interactive exchanges and individual work on laptops. 
Workspaces must also accommodate interactive project work where small teams can gather and work together uninterrupted by other groups.    
Universities are an obvious first choice for hosting a hack week given their access to scientific resources, research support and infrastructure. 
However traditional university lecture halls often do not meet the hack week criteria for interactive exchanges.
Fortunately active learning teaching methods are creating new environments that foster interactivity through modified seating arrangements and integration of computer monitors at group workstations. There is a natural tension between keeping the group together and providing physically separated break-out spaces. On the one hand, conducting the workshop in a single large venue improves group cohesion and prevents self-segregation of researchers and ideas. On the other hand, the interactive nature of the workshop may lead to an environment where it is difficult for individuals to focus on the highly complex tasks that are typical for hack week projects. It may hence be advisable to allow for diffusion of the group into adjacent rooms, while providing ample venue for the entire group to congregate for tutorials, breaks and reports.

>>>>

A hackweek requires a flexible workspace environment with ample opportunity for re-configuration. 
Participants must have access to rooms that can accommodate lectures combined with interactive exchanges and individual work on laptops. 
Workspaces must also accommodate interactive project work where small teams can gather and work together uninterrupted by other groups.    
Universities are an obvious first choice for hosting a hack week given their access to scientific resources, research support and infrastructure. 
However traditional university lecture halls often do not meet the hack week criteria for interactive exchanges.
Fortunately, many universities are experimenting with other types of spaces: for example, the adoption of active learning teaching methods \cite{prince2004} has led to the development of modular classrooms, where seating arrangements can be flexibly modified and group activities can be more readily undertaken. There is a natural tension between keeping the group together and providing physically separated break-out spaces. On the one hand, conducting the workshop in a single large venue improves group cohesion and prevents self-segregation of researchers and ideas. On the other hand, the interactive nature of the workshop may lead to an environment where it is difficult for individuals to focus on the highly complex tasks that are typical for hack week projects. It may hence be advisable to allow for diffusion of the group into adjacent rooms, while providing ample venue for the entire group to congregate for tutorials, breaks and reports.

### 2017-09-30 sections/intro.tex
Added Figure 1 back, fixed links to figures

%\section*{}
\label{sec:introduction}

As data becomes cheaper to gather and store, research across a wide range of disciplines has become increasingly reliant on computational workflows involving a familiarity with aspects of statistical modeling, machine learning, scalable computation, and related skills. In addition, the recent reproducibility crises in several scientific fields has led to the growing realization that improving awareness of open science and reproducibility as well as practical skills in making research reproducible is essential to scientific progress \cite[e.g.][]{pashler2012,frye2015,gezelter2015,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops (an overview and typography of such efforts in the data science context can be found in \cite{demasi2017}).
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program  \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of domain-specific courses focusing on statistics and computation within their field.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons that take place in conjunction with meetings of the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}.
Generally, pedagogically-focused events follow a classic academic model where novices learn new skills from experts, while project-focused workshops emphasize collaborative activities using existing skills.
A disadvantage of the pedagogical model is that it can tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the project model is the common perception that the week is designed for technical experts, which may discourage others from attending.
In 2014, we initiated an alternative model of ``Hack Weeks'' that try to fill the gaps between these models.
These are week-long events that combine pedagogy (often focused on statistical and computational techniques) together with time for hacks and creative projects, and with the goal of encouraging collaboration and learning among people at various stages of their career.

%\begin{figure}
%\begin{center}
%\includegraphics[width=9cm]{fig/HackSpectrum}
%\caption{Comparison of Extracurricular Workshop Models}
%\label{fig:hackspectrum}
%\end{center}
%\end{figure}

>>>>

%\section*{}
\label{sec:introduction}
\dropcap{A}s data becomes cheaper to gather and store, research across a wide range of disciplines has become increasingly reliant on computational workflows involving a familiarity with aspects of statistical modeling, machine learning, scalable computation, and related skills. In addition, the recent reproducibility crises in several scientific fields has led to the growing realization that improving awareness of open science and reproducibility as well as practical skills in making research reproducible is essential to scientific progress \cite[e.g.][]{pashler2012,frye2015,gezelter2015,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops (an overview and typography of such efforts in the data science context can be found in \cite{demasi2017}).
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program  \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of domain-specific courses focusing on statistics and computation within their field.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons that take place in conjunction with meetings of the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}.
Generally, pedagogically-focused events follow a classic academic model where novices learn new skills from experts, while project-focused workshops emphasize collaborative activities using existing skills.
A disadvantage of the pedagogical model is that it can tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the project model is the common perception that the week is designed for technical experts, which may discourage others from attending.
In 2014, we initiated an alternative model of ``Hack Weeks'' that try to fill the gaps between these models.
These are week-long events that combine pedagogy (often focused on statistical and computational techniques) together with time for hacks and creative projects, and with the goal of encouraging collaboration and learning among people at various stages of their career.

\begin{figure}
\begin{center}
\includegraphics[width=9cm]{fig/HackSpectrum}
\caption{Comparison of Extracurricular Workshop Models}
\label{fig:hackspectrum}
\end{center}
\end{figure}

### 2017-09-30 sections/results.tex

Measuring the success of a hack week objectively is complicated by the variety of objectives that a hack week might have (see above).
Additionally, the participant-driven, open-ended format facilitates knowledge transfer and collaborations in sometimes surprising ways that escape traditional measures of success.
One key metric is the number of publications that result from hack week projects, but this is a fairly narrow definition of success, in line with standard academic performance indicators.
Assuming that participants work largely in the open during a hack week, and that most projects have a strong programming component another indicator of success is the activity of participants in terms of code written and committed to a public code repository.
Still, these measures ignore learning, community-building as well as networking outcomes.
These outcomes can be assessed through post-workshop surveys.
%It is in principle possible to measure gains in knowledge, networks and productivity within the pool of acceptable candidates for both those that attended the hack week and those that did not.
%This would then provide a somewhat objective measure of the impact the hack week has had.
%In practice, this has not been done for any of the hack weeks conducted thus far.
Here, we have taken an approach that combines these metrics: we start with survey results, and anecdotally report about publications, new code and projects generated (see \ref{box:outcomes})
%Open-ended questions allow participants to provide feedback about outcomes, problems and goals that organizers had not anticipated.

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks from 2016, we find that all participants self-report successful learning outcomes in new topics, tools or methods (AHW: 86\%, GHW: 94\%, NHW: 76\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, top panel, left).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist ( Figure \ref{fig:survey}, top panel, right and middle).
%While each hack week also probed participants' attitudes in more detail toward specific topics, methods, and modes of learning, the hack weeks differed substantially on that level, making the results difficult to compare.
%Another important goal of the hack weeks was centred on community building and fostering collaborations.
The majority of participants felt that their contributions to their hack teams was valued, and that they built valuable connections to other researchers (Figure \ref{fig:survey}, middle panel, middle).
This is especially true for Neuro Hack Week, where more than 64\% of participants strongly agreed that they formed valuable connections at NHW (Figure  \ref{fig:survey}, middle panel, right).
Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 83\%, GHW: 69\%, NHW: 71\% for combined responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, middle panel, left), though responses are not as unequivocal as they are in some of the other categories.
This is expected: participants new to the field may participate to learn rather than to teach.

We find that the hack weeks have been largely successful at efforts to promote positive attitudes towards reproducibility and open science: at all three events, more than 70\% of all participants (AHW: 79\%, GHW: 72\%, NHW:85\%; Figure \ref{fig:survey}, bottom panel, left) put code or data created at the hack week into a public repository, while a substantially smaller fraction of participants followed this practice before the event (Figure \ref{fig:survey}, bottom panel, middle).
When asking participants whether they had made their code or data openly available in the past, the overall behaviour reflects how general conventions and attitudes differ in different fields.
In particular, astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%Neuroscience seems to fall somewhere in between.
This implies that hack weeks can have the highest impacts in field where a priori engagement in reproducibility efforts is low and significant progress can be made towards changing researchers' attitudes during a collaborative workshop.
Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science: again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, closely followed by neuroscience (95\%), while there was somewhat less of an impact on participants' attitudes in astronomy (72\%).
Overall, our results show that hack weeks are effective at addressing persisting doubts about making research open and reproducible.

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=12cm]{fig/eval_techskills}
\includegraphics[width=12cm]{fig/eval_collab}
\includegraphics[width=12cm]{fig/eval_openscience}
\caption{{\bf Post-workshop surveys from three hack weeks}: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. We report here about results in three different domains: the development of technical skills (top), collaboration and learning (middle), and shifts in attitudes towards reproducibility and open science (bottom)}
\label{fig:survey}
\end{center}
\end{figure*}

@@

\subsection*{Examples of Hack Week Outcomes}
\label{sec:outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used the opportunity of Astro Hack Week to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing well-tested, well-documented implementations of algorithms for time series analysis often used in X-ray astronomy.
The start of this project was facilitated by the collaborative environment at Astro Hack Week, including expertise in how to start/run open-source projects, role models of successful projects, and an environment encouraging scientific risk taking. Astro Hack Week enabled participants to seed a new collaboration around a software project needed by the larger community.
Since its beginnings at Astro Hack Week, Stingray has matured into an enduring collaboration within the community with five active maintainers, a number of contributors and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week}
In 2016, a Geo Hack Week project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States.
The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine\footnote{\url{http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/}}.
George Richardson, one of the project leads, now works for a renewable resource company in Seattle.
\subsubsection*{Example 3: Neuro Hack Week}
Motion of study participants inside of the MRI machine is a major concern in neuroimaging studies.
This is a particular concern in studies in which children or patients are studied, as they are more likely to move.
In the 2016 Neuro Hack Week one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off were varied.
They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four different researchers from four different institutions in two different countries) continued to work on this project remotely after the Neuro Hack Week was over, and they eventually wrote and published a paper describing these results in the open access journal Research Ideas and Outcomes \cite{leonard2017}.

>>>>

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=11cm]{fig/eval_techskills}
\includegraphics[width=11cm]{fig/eval_collab}
\includegraphics[width=11cm]{fig/eval_openscience}
\caption{{\bf Post-workshop surveys from three hack weeks}: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. We report here about results in three different domains: the development of technical skills (top), collaboration and learning (middle), and shifts in attitudes towards reproducibility and open science (bottom)}
\label{fig:survey}
\end{center}
\end{figure*}

Measuring the success of a hack week objectively is complicated by the variety of objectives that a hack week might have (see above). 
Additionally, the participant-driven, open-ended format facilitates knowledge transfer and collaborations in sometimes surprising ways that escape traditional measures of success.

One key metric is the number of publications that result from hack week projects, but this is a fairly narrow definition of success, in line with standard academic performance indicators.
Assuming that participants work largely in the open during a hack week, and that most projects have a strong programming component another indicator of success is the activity of participants in terms of code written and committed to a public code repository.
Still, these measures ignore learning, community-building as well as networking outcomes, which can be assessed through post-workshop surveys.
%It is in principle possible to measure gains in knowledge, networks and productivity within the pool of acceptable candidates for both those that attended the hack week and those that did not.
%This would then provide a somewhat objective measure of the impact the hack week has had.
%In practice, this has not been done for any of the hack weeks conducted thus far.
Here, we have taken an approach that combines these metrics: we start with survey results, and anecdotally report about publications, new code and projects generated (see the following section).
%Open-ended questions allow participants to provide feedback about outcomes, problems and goals that organizers had not anticipated.

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks (AHW, GHW, NHW, respectively) from 2016, we find that all participants self-report successful learning outcomes in new topics, tools or methods (AHW: 86\%, GHW: 94\%, NHW: 76\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, top panel, left).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist ( Figure \ref{fig:survey}, top panel, right and middle).
%While each hack week also probed participants' attitudes in more detail toward specific topics, methods, and modes of learning, the hack weeks differed substantially on that level, making the results difficult to compare.
%Another important goal of the hack weeks was centred on community building and fostering collaborations.
The majority of participants felt that their contributions to their hack teams was valued, and that they built valuable connections to other researchers (Figure \ref{fig:survey}, middle panel, middle).
This is especially true for Neuro Hack Week, where more than 64\% of participants strongly agreed that they formed valuable connections at NHW (Figure  \ref{fig:survey}, middle panel, right).
Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 83\%, GHW: 69\%, NHW: 71\% for combined responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, middle panel, left), though responses are not as unequivocal as they are in some of the other categories.
This is expected: participants new to the field may participate to learn rather than to teach.

We find that the hack weeks have been largely successful at efforts to promote positive attitudes towards reproducibility and open science: at all three events, more than 70\% of all participants (AHW: 79\%, GHW: 72\%, NHW:85\%; Figure \ref{fig:survey}, bottom panel, left) put code or data created at the hack week into a public repository, while a substantially smaller fraction of participants followed this practice before the event (Figure \ref{fig:survey}, bottom panel, middle).
When asking participants whether they had made their code or data openly available in the past, the overall behaviour reflects how general conventions and attitudes differ in different fields.
Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
This implies that hack weeks can have the highest impacts in field where a priori engagement in reproducibility efforts is low and significant progress can be made towards changing researchers' attitudes during a collaborative workshop.
Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science: again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, closely followed by neuroscience (95\%), while there was somewhat less of an impact on participants' attitudes in astronomy (72\%).
Overall, our results show that hack weeks are effective at addressing persisting doubts about making research open and reproducible.

@@

\subsection*{Examples of Hack Week Outcomes}
\label{sec:outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used the opportunity of AHW to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing well-tested, well-documented implementations of time series analysis algorithms often used in X-ray astronomy.
The start of this project was facilitated by the collaborative environment at Astro Hack Week, including expertise in how to start/run open-source projects, role models of successful projects, and an environment encouraging scientific risk taking. Astro Hack Week enabled participants to seed a new collaboration around a software project needed by the larger community.
Since its beginnings at Astro Hack Week, Stingray has matured into an enduring collaboration within the community with five active maintainers, a number of contributors and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week}
In 2016, a GHW project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States.
The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine\footnote{\url{http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/}}.
George Richardson, one of the project leads, now works for a renewable resource company in Seattle.
\subsubsection*{Example 3: Neuro Hack Week}
Motion of study participants inside of the MRI machine is a major concern in neuroimaging studies, particularly in studies of children or patients, as they are more likely to move.
During NHW 2016 one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off were varied.
%They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four different researchers from four different institutions in two different countries) continued to work on this project remotely after the end of Neuro Hack Week, and eventually published a paper describing these results in the open access journal Research Ideas and Outcomes \cite{leonard2017}.
